---
title: "fordgobikes"
author: "jagger241"
date: "2/3/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

In this notebook I download and unzip the [FordGoBike](https://www.fordgobike.com/) [data](https://www.fordgobike.com/system-data). 

```{r}

library(pacman)
p_load(tidyverse, tictoc, ggmap, skimr, lubridate, forcats, biganalytics, doParallel)

```

Downloading the data directly from https://s3.amazonaws.com/fordgobike-data 

```{r, eval=FALSE}

#2017 includes all months into a single file

URL <- "https://s3.amazonaws.com/fordgobike-data/2017-fordgobike-tripdata.csv"
download.file(URL, destfile = "./data/2017-fordgobike-tripdata.csv", method="curl")

#in 2018, data was seperated by month so I loop over to download everything from January to Jully

for (i in 1:7) {
URL <- paste0("https://s3.amazonaws.com/fordgobike-data/20180",i,"-fordgobike-tripdata.csv.zip")
download.file(URL, destfile = paste0("./data/20180",i,"-fordgobike-tripdata.csv.zip"), method="curl")
}
```

Unzip downloaded files.

```{r}
unzip("./data/201801-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201802-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201803-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201804-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201805-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201806-fordgobike-tripdata.csv.zip",exdir="./data")
unzip("./data/201807-fordgobike-tripdata.csv.zip",exdir="./data")
```

Clean up data directory.

```{r}
fn <- "./data/201801-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201802-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201803-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201804-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201805-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201806-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
fn <- "./data/201807-fordgobike-tripdata.csv.zip"
if (file.exists(fn)) file.remove(fn)
```

Read the.csv files

```{r message=FALSE}
fordgobike2017 <- read_csv(file="./data/2017-fordgobike-tripdata.csv")
fordgobike201801 <- read_csv(file="./data/201801-fordgobike-tripdata.csv")
fordgobike201802 <- read_csv(file="./data/201802-fordgobike-tripdata.csv")
fordgobike201803 <- read_csv(file="./data/201803-fordgobike-tripdata.csv")
fordgobike201804 <- read_csv(file="./data/201804-fordgobike-tripdata.csv")
fordgobike201805 <- read_csv(file="./data/201805-fordgobike-tripdata.csv")
fordgobike201806 <- read_csv(file="./data/201806-fordgobike-tripdata.csv")
fordgobike201807 <- read_csv(file="./data/201807-fordgobike-tripdata.csv")
```

Check the head() of the loaded data.frames 


```{r}
head(fordgobike2017,3) 
head(fordgobike201801,3)
head(fordgobike201802,3)
head(fordgobike201803,3)
head(fordgobike201804,3)
head(fordgobike201805,3)
head(fordgobike201806,3)
head(fordgobike201807,3)
```
Check the tail() of the loaded data.frames.

```{r}
tail(fordgobike2017,3)
tail(fordgobike201801,3)
tail(fordgobike201802,3)
tail(fordgobike201803,3)
tail(fordgobike201804,3)
tail(fordgobike201805,3)
tail(fordgobike201806,3)
tail(fordgobike201807,3)
```

Check the dimension (number of rows and columns) of the data
```{r}

dim(fordgobike2017)
dim(fordgobike201801)
dim(fordgobike201802)
dim(fordgobike201803)
dim(fordgobike201804)
dim(fordgobike201805)
dim(fordgobike201806)
dim(fordgobike201807)
```

Change the data types in 2017 variables to be uniform and then merge everything

```{r}

fordgobike201806 <- fordgobike201806 %>%
        mutate(start_station_id = as.integer(start_station_id),
               end_station_id= as.integer(end_station_id) )   

fordgobike201807 <- fordgobike201807 %>%
        mutate(start_station_id = as.integer(start_station_id),
               end_station_id= as.integer(end_station_id) )

fordgobike2018 <- bind_rows(fordgobike201801, fordgobike201802, fordgobike201803, 
                            fordgobike201804,fordgobike201805, fordgobike201806,
                            fordgobike201807)

glimpse(fordgobike2018)

```

Merge 2017 and 2018 data
```{r}
#check the dimension first
dim(fordgobike2017)
dim(fordgobike2018)

fordgobike <- bind_rows(fordgobike2017, fordgobike2018)
head(fordgobike, 3)

#output the data as a csv file
write.csv(fordgobike, file = "./data/fordgobike.csv")

dim(fordgobike)

```

create new variable for "age", "year", "month", and "day"

```{r}

fordgobike <- fordgobike %>% mutate(age = year(now()) - member_birth_year, year=year(start_time), month=month(start_time), day=day(start_time))

head(fordgobike,3)

dim(fordgobike)

```

Create a new variable "weekday"
```{r}

fordgobike <- fordgobike %>% mutate(week_day = wday(start_time) )

levels <- c("M","T","W","TH","F","SAT","SUN")

fordgobike$week_day <- factor(fordgobike$week_day, levels = levels)

head(fordgobike, 3)
dim(fordgobike)
```

Distribution of riders by "age", filter riders below 81, 101, and above 100.

```{r warning=FALSE}

fordgobike %>% group_by(age) %>% count()

fordgobike %>% group_by(age) %>% summary()

skim(fordgobike)

fordgobike %>% ggplot(aes(x=age)) + geom_histogram()

fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age)) + geom_histogram()

fordgobike %>% filter(age <= 100) %>% ggplot(aes(x=age)) + geom_histogram()

fordgobike %>% filter(age > 100) %>% ggplot(aes(x=age)) + geom_histogram()
```

Group by "gender", "age" (below 81) and plot their histograms
```{r}
fordgobike %>% group_by( member_gender, age ) %>% count()

fordgobike %>% ggplot(aes(x=age, class=member_gender)) + geom_histogram()

fordgobike %>% ggplot(aes(x=age, class=member_gender)) + geom_histogram(aes(y=..density..))

fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age)) + geom_histogram()
```

Plot histograms of people below 81 years old and facet by gender with relative colors
```{r}
fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age, color=member_gender)) + 
  geom_histogram(position="identity") + 
  facet_grid(member_gender ~ .)

fordgobike %>% filter(age <= 80) %>% ggplot(aes(x=age, color=member_gender)) + 
  geom_histogram(aes(y=..density..),position="identity") + 
  facet_grid(member_gender ~ .)
```

Plotting the "Year", "month", and day of week.

```{r}

fordgobike %>% ggplot(aes(x=year)) + geom_bar()
fordgobike %>% ggplot(aes(x=month)) + geom_bar() + facet_grid(year ~ .)
fordgobike %>% ggplot(aes(x=day)) + geom_bar() + facet_grid(year ~ .)
```


Removing geocode outliers, subset longitude and latitude, and plot the riders' location in the Bay Area
```{r}
fordgobike2018 <- fordgobike2018 %>% filter(start_station_latitude < 38 & start_station_longitude < 120 )

fordgobike_subset <- fordgobike2018 %>% select(start_station_longitude,start_station_latitude)

fordgobike_subset %>% ggplot(aes(x=start_station_longitude, y=start_station_latitude)) +
  geom_point()
```

store the data as matrix, create a k-means cluster for each location (Oakland, San Jose, San francisco), and map them  

```{r}
tic()
registerDoParallel(cores = 8)

head(fordgobike2018)

fordgobike_subset2 <- as.matrix(fordgobike_subset)

set.seed <- 123454321

cl <- bigkmeans(fordgobike_subset2, 3, nstart=8)

cl$centers

fordgobike2018 <- fordgobike2018 %>% 
  mutate(clust = cl$cluster)

fordgobike_subset %>% ggplot(aes(x=start_station_longitude, y=start_station_latitude, color=cl$cluster)) +
  geom_point()

toc()
```

Plotting the stations 
```{r}
# https://stackoverflow.com/questions/20621250/simple-approach-to-assigning-clusters-for-new-data-after-k-means-clustering

cl$centers

closest.cluster <- function(x) {
  cluster.dist <- apply(cl$centers, 1, function(y) sqrt(sum((x-y)^2)))
  return(which.min(cluster.dist)[1])
}

oak <- closest.cluster(c(-122.2711, 37.8044))
sj <- closest.cluster(c(-121.8953, 37.34168))
sf <- closest.cluster(c(-122.4072, 37.77809))

#Oakland stations
oakland <- fordgobike2018 %>% filter(clust == oak) 
oakland %>% ggplot(aes(x=start_station_longitude, y=start_station_latitude)) +
  geom_point() +
  ggtitle("Oakland Ford Go Bike stations")

#San Jose  stations
san_jose <- fordgobike2018 %>% filter(clust == sj) 
san_jose %>% ggplot(aes(x=start_station_longitude, y=start_station_latitude)) +
  geom_point() +
  ggtitle("San Jose Ford Go Bike stations")

#San Francisco stations
san_francisco <- fordgobike2018 %>% filter(clust == sf) 
san_francisco %>% ggplot(aes(x=start_station_longitude, y=start_station_latitude)) +
  geom_point() +
  ggtitle("San Francisco Ford Go Bike stations")
```


Register with Google Maps and locate bike stations 
```{r}
#register_google(key = "XXXXXXXXXXXXXXXXX-XXXXXXXXXX", write = TRUE)

#Oakland
get_map(location = c(lon=cl$centers[oak,1], lat=cl$centers[oak,2]), zoom = 12, maptype = "roadmap") %>% ggmap() +
  geom_point(data = oakland, aes(x = start_station_longitude, y = start_station_latitude), size = 1, shape = 19) +
  ggtitle("Oakland Ford Go Bike stations")

#San Jose
get_map(location = c(lon=cl$centers[sj,1], lat=cl$centers[sj,2]), zoom = 12, maptype = "roadmap") %>% ggmap() +
  geom_point(data = san_jose, aes(x = start_station_longitude, y = start_station_latitude), size = 1, shape = 19) +
  ggtitle("San Jose Ford Go Bike stations")

#San Francisco
get_map(location = c(lon=cl$centers[sf,1], lat=cl$centers[sf,2]), zoom = 12, maptype = "roadmap") %>% ggmap() +
  geom_point(data = san_francisco, aes(x = start_station_longitude, y = start_station_latitude), size = 1, shape = 19) +
  ggtitle("San Francisco Ford Go Bike stations")

```

Bike stations in the Bay Area
```{r}
tic()

#I chose hayward to better capture San Jose
bayarea <- get_map(location = "hayward") 

ggmap(bayarea) +
  geom_point(data = fordgobike2018, aes(x = start_station_longitude, y = start_station_latitude, color  = clust, alpha = 0.1), size = 1, shape = 19) +
  ggtitle("Bay Area Ford Go Bike stations")
toc()
```

Gender of users in the Bay Area, and then by city (Oakland = 1, San Jose = 2, San Francisco = 3)

```{r}
#Duration Distribution by Gender
fordgobike2018 %>% ggplot(aes(x = member_gender, y = duration_sec/1000000)) + geom_bar(stat = "Identity") +
  ylab("Duration (in Million Sec)") +
  xlab("Gender") +
  ggtitle("Duration Distribution By Gender")

#Duration Distribution by Gender for each city
fordgobike2018 %>% 
  mutate(clust = 
           ifelse(clust %in% closest.cluster(c(-122.2711, 37.8044)),
                  "Oakland", 
                  ifelse(clust %in% closest.cluster(c(-121.8953, 37.34168)), "San Jose", 
                        "San Francisco"))) %>%
  ggplot(aes(x=member_gender, y=duration_sec/1000000)) +
  geom_bar(stat="Identity") +
  ggtitle("Bay Area") +
  ylab("Duration (in Million Sec)") +
  xlab("Gender") +
  ggtitle("Duration Distribution by Gender for Each City") +
  facet_grid(clust ~ .)

```

Plot the density histograms of ride durations in the Bay Area by gender

```{r}
#density histograms of ride durations
fordgobike2018 %>% ggplot(aes(x=duration_sec, y=..density..)) + 
  scale_x_continuous(limits = c(0, 10000)) +
  geom_histogram() +
  geom_density(aes(y=..density..)) 

#density histograms of ride durations logged
fordgobike2018 %>% ggplot(aes(log(x=duration_sec), y=..density..)) +
  geom_histogram() +
  geom_density(aes(y=..density..)) 

#density histograms of ride durations by gender
fordgobike2018 %>% ggplot(aes(x=duration_sec, y=..density..)) + 
  scale_x_continuous(limits = c(0, 10000)) +
  geom_histogram() + 
  geom_density(aes(y=..density..)) +
  facet_grid(member_gender ~ .)

#density histograms of ride durations by gender logged
fordgobike2018 %>% ggplot(aes(log(x=duration_sec), y=..density..)) +
  geom_histogram() + 
  geom_density(aes(y=..density..)) +
  facet_grid(member_gender ~ .)
```

summary by gender for each city

```{r}
#Oak
fordgobike2018 %>% filter(clust == 1) %>% 
  group_by( member_gender ) %>%
  summarize(dur_mean = mean(duration_sec), dur_sd = sd(duration_sec))

#SJ
fordgobike2018 %>% filter(clust == 2) %>% 
  group_by( member_gender ) %>%
  summarize(dur_mean = mean(duration_sec), dur_sd = sd(duration_sec))

#SF
fordgobike2018 %>% filter(clust == 3) %>% 
  group_by( member_gender ) %>%
  summarize(dur_mean = mean(duration_sec), dur_sd = sd(duration_sec))
```

